---
title: "LUREC_research_code"
author: "Justin Peabody"
date: "2026-04-30"
output: md_document
---

```{r}

library(tidyverse)
library(plotly)
library(regclass)

```
# Data cleaning, exploratory analysis, and visualization

The very first action taken, before the project began, was downloading nearby precipitation data from this site: <https://www.ncei.noaa.gov/access/past-weather/Woodstock%2C%20Illinois.>

This was later visualized against the soil water content data from the field site. 

```{r, echo=FALSE}

Woodstock_0.7_SW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_0.7_SW_weather_data.csv")
Woodstock_3.8_SW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_3.8_SW_data.csv")
Woodstock_5_NW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_5_NW_data.csv")

```
I then prepared the data for visualization and analysis by reformatting each date field to date class, filtering rows for 2025, and combining each station's data into a single dataset.

```{r}

Woodstock_0.7_SW_data<-Woodstock_0.7_SW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))

Woodstock_3.8_SW_data<-Woodstock_3.8_SW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))
  
Woodstock_5_NW_data<-Woodstock_5_NW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))
  
Woodstock_combined_weather_data<-Woodstock_0.7_SW_data%>%
  left_join(Woodstock_3.8_SW_data, by=c("Date"="Date"))%>%
  left_join(Woodstock_5_NW_data, by=c("Date"="Date"))%>%
  select(1,2,6,10)

names(Woodstock_combined_weather_data)<-c("Date","SW_0.7_Precip", "SW_3.8_Precip", "NW_5_Precip")

```

At the onset of my Environmental Research class in January, I obtained the onsite soil moisture data from my professor, which includes moisture data from November 10, 2024 (just after being installed) to October 6, 2025.

```{r}
six_LUREC_water_sensor_data<-read_csv("~/Desktop/Ghatak_research_project/6_LUREC_soil_water_sensors.csv")
six__more_LUREC_water_sensors<-read_csv("~/Desktop/Ghatak_research_project/6_more_LUREC_water_sensors.csv")

```
Each dataset contains readings from a single data logger, which receives data from six of the twelve sensors (2 datasets = 2 data loggers = 12 total sensors). 

I then performed some basic data cleaning, including removal of column headers that were formatted as observations and renaming columns to prevent word splitting issues, and changing water content readings from character to numeric class.

```{r}

six_LUREC_water_sensor_data_new<-six_LUREC_water_sensor_data%>%
tail(7916)
colnames(six_LUREC_water_sensor_data_new)<-c("Timestamp", "Pot_1_Water_Content_proportion", "Pot_2_Water_Content_proportion", "Pot_3_Water_Content_proportion", "Pot_4_Water_Content_proportion", "Pot_5_Water_Content_proportion", "Pot_6_Water_Content_proportion", "Battery_percent", "Battery_voltage_mV")

six__more_LUREC_water_sensors_new<-six__more_LUREC_water_sensors%>%
tail(7910)
colnames(six__more_LUREC_water_sensors_new)<-c("Timestamp", "Pot_7_Water_Content_proportion", "Pot_8_Water_Content_proportion", "Pot_9_Water_Content_proportion", "Pot_10_Water_Content_proportion", "Pot_11_Water_Content_proportion", "Pot_12_Water_Content_proportion", "Battery_percent", "Battery_voltage_mV")

rm(six_LUREC_water_sensor_data, six__more_LUREC_water_sensors)

six_LUREC_water_sensor_data_new<-six_LUREC_water_sensor_data_new%>%
mutate(Pot_1_Water_Content_proportion=as.numeric(Pot_1_Water_Content_proportion))%>%
mutate(Pot_2_Water_Content_proportion=as.numeric(Pot_2_Water_Content_proportion))%>%
mutate(Pot_3_Water_Content_proportion=as.numeric(Pot_3_Water_Content_proportion))%>%
mutate(Pot_4_Water_Content_proportion=as.numeric(Pot_4_Water_Content_proportion))%>%
mutate(Pot_5_Water_Content_proportion=as.numeric(Pot_5_Water_Content_proportion))%>%
mutate(Pot_6_Water_Content_proportion=as.numeric(Pot_6_Water_Content_proportion))

six_more_LUREC_water_sensors_new<-six__more_LUREC_water_sensors_new%>%
  mutate(Pot_7_Water_Content_proportion=as.numeric(Pot_7_Water_Content_proportion))%>%
  mutate(Pot_8_Water_Content_proportion=as.numeric(Pot_8_Water_Content_proportion))%>%
  mutate(Pot_9_Water_Content_proportion=as.numeric(Pot_9_Water_Content_proportion))%>%
  mutate(Pot_10_Water_Content_proportion=as.numeric(Pot_10_Water_Content_proportion))%>%
  mutate(Pot_11_Water_Content_proportion=as.numeric(Pot_11_Water_Content_proportion))%>%
  mutate(Pot_12_Water_Content_proportion=as.numeric(Pot_12_Water_Content_proportion))

rm(six__more_LUREC_water_sensors_new)

six_LUREC_water_sensor_data_new<-six_LUREC_water_sensor_data_new%>%
tail(7915)
Combined_LUREC_water_data<-six_LUREC_water_sensor_data_new%>%
left_join(six_more_LUREC_water_sensors_new, by=c("Timestamp"="Timestamp"))

```

I then investigated the distribution of each pot's water content.

```{r}

Averaged_LUREC_water_data<-Combined_LUREC_water_data%>%
mutate(Timestamp=mdy_hm(Timestamp))%>%
mutate(Day=format(Timestamp, format= "%d"), Month=format(Timestamp, format="%m"), Year=format(Timestamp, format="%y"))%>%
mutate(Date=str_c(Month, "/", Day, "/", Year))%>%
mutate(Date=mdy(Date))%>%
group_by(Date)%>%
summarize(Pot_1_mean_water_content=mean(Pot_1_Water_Content_proportion), Pot_2_mean_water_content=mean(Pot_2_Water_Content_proportion), Pot_3_mean_water_content=mean(Pot_3_Water_Content_proportion), Pot_4_mean_water_content=mean(Pot_4_Water_Content_proportion), Pot_5_mean_water_content=mean(Pot_5_Water_Content_proportion), Pot_6_mean_water_content=mean(Pot_6_Water_Content_proportion), Pot_7_mean_water_content=mean(Pot_7_Water_Content_proportion), Pot_8_mean_water_content=mean(Pot_8_Water_Content_proportion), Pot_9_mean_water_content=mean(Pot_9_Water_Content_proportion), Pot_10_mean_water_content=mean(Pot_10_Water_Content_proportion), Pot_11_mean_water_content=mean(Pot_11_Water_Content_proportion), Pot_12_mean_water_content=mean(Pot_12_Water_Content_proportion))    

Water_content_precipitation_combined<-Averaged_LUREC_water_data%>%
left_join(Woodstock_combined_weather_data, by=c("Date"="Date"))%>%
mutate(Avg_MWC=(Pot_1_mean_water_content+Pot_2_mean_water_content+Pot_3_mean_water_content+Pot_4_mean_water_content+Pot_5_mean_water_content+Pot_6_mean_water_content+Pot_7_mean_water_content+Pot_8_mean_water_content+Pot_9_mean_water_content+Pot_10_mean_water_content+Pot_11_mean_water_content+Pot_12_mean_water_content)/12)%>%
select(-17)

colnames(Water_content_precipitation_combined)=c("Date", "Pot_1_MWC", "Pot_2_MWC", "Pot_3_MWC", "Pot_4_MWC", "Pot_5_MWC", "Pot_6_MWC", "Pot_7_MWC", "Pot_8_MWC", "Pot_9_MWC", "Pot_10_MWC", "Pot_11_MWC", "Pot_12_MWC","PRCP_x", "PRCP_y", "PRCP_z")
Water_content_precipitation_combined<-Water_content_precipitation_combined%>%
mutate(Avg_precip=(PRCP_x+PRCP_y+PRCP_z)/3)

Combined_LUREC_water_data<-six_LUREC_water_sensor_data_new%>%
tail(7915)%>%
left_join(six_more_LUREC_water_sensors_new, by=c("Timestamp"="Timestamp"))

Combined_LUREC_water_data_pivoted<-Combined_LUREC_water_data%>%
select(-Battery_percent.x, -Battery_voltage_mV.x, -Battery_percent.y, -Battery_voltage_mV.y)%>%
pivot_longer(cols=-Timestamp, names_to="Pot", values_to="Water_content")

Combined_LUREC_water_data_pivoted$Pot<-rep(1:12, 7915)
Combined_LUREC_water_data_pivoted$Pot<-as.factor(Combined_LUREC_water_data_pivoted$Pot)

Pot_water_content_boxplots<-plot_ly(data=Combined_LUREC_water_data_pivoted, color=~Pot, y=~Water_content, type="box")%>%
layout(title="Comparative boxplots of Water Content for Each LUREC Pot", xaxis=list(title="Pot Number"), yaxis=list(title="Water content (m^3 water/m^3 soil"))
Pot_water_content_boxplots

```

Creating this boxplot made it visually apparent that pots 5 and 8 contained negative values. Since soil moisture content is measured as a proportion of cubic meters of water per cubic meter of soil, negative moisture readings make no sense. 

I then investigated when these negative readings occured.

```{r}

Pot_5_negative_values<-Combined_LUREC_water_data%>%
select(Timestamp, Pot_5_Water_Content_proportion)%>%
filter(Pot_5_Water_Content_proportion<0)

Pot_8_negative_values<-Combined_LUREC_water_data%>%
select(Timestamp, Pot_8_Water_Content_proportion)%>%
filter(Pot_8_Water_Content_proportion<0)

```

The sensor in Pot 5 recorded negative values for water content at each hour mark from 1/21/25 at 9:00 AM to 1/22/25 at 7:00 PM. The sensor in Pot 8 recorded negative values for water content at each hour mark from 10:00 AM to 5:00 PM on 1/21/26. The negative values then discontinue and begin again from 12:00 AM to 5:00 PM on 1/22/25.

Based on this information and the locations of large clusters of NA values between the months of November and January for 2 of the pots, I decided to remove all observations at or before 2/3/25, after averaging the observations by day. This was the dataset used for all subsequent analyses.

```{r}

Averaged_LUREC_water_data<-Combined_LUREC_water_data%>%
mutate(Timestamp=mdy_hm(Timestamp))%>%
mutate(Day=format(Timestamp, format= "%d"), Month=format(Timestamp, format="%m"), Year=format(Timestamp, format="%y"))%>%
mutate(Date=str_c(Month, "/", Day, "/", Year))%>%
mutate(Date=mdy(Date))%>%
group_by(Date)%>%
summarize(Pot_1_mean_water_content=mean(Pot_1_Water_Content_proportion), Pot_2_mean_water_content=mean(Pot_2_Water_Content_proportion), Pot_3_mean_water_content=mean(Pot_3_Water_Content_proportion), Pot_4_mean_water_content=mean(Pot_4_Water_Content_proportion), Pot_5_mean_water_content=mean(Pot_5_Water_Content_proportion), Pot_6_mean_water_content=mean(Pot_6_Water_Content_proportion), Pot_7_mean_water_content=mean(Pot_7_Water_Content_proportion), Pot_8_mean_water_content=mean(Pot_8_Water_Content_proportion), Pot_9_mean_water_content=mean(Pot_9_Water_Content_proportion), Pot_10_mean_water_content=mean(Pot_10_Water_Content_proportion), Pot_11_mean_water_content=mean(Pot_11_Water_Content_proportion), Pot_12_mean_water_content=mean(Pot_12_Water_Content_proportion)) 

Water_content_precipitation_combined<-Averaged_LUREC_water_data%>%
left_join(Woodstock_combined_weather_data, by=c("Date"="Date"))%>%
mutate(Avg_MWC=(Pot_1_mean_water_content+Pot_2_mean_water_content+Pot_3_mean_water_content+Pot_4_mean_water_content+Pot_5_mean_water_content+Pot_6_mean_water_content+Pot_7_mean_water_content+Pot_8_mean_water_content+Pot_9_mean_water_content+Pot_10_mean_water_content+Pot_11_mean_water_content+Pot_12_mean_water_content)/12)

Useful_LUREC_averaged_data<-Water_content_precipitation_combined%>%
filter(Date > "2025-02-03")

```

I then created a three-axis time series graph displaying daily averaged water content against the daily precipitation and maximum temperature at the Crystal Lake 4 NW weather station (the only weather station near the field site with both temperature and precipitation data for 2025). Data for this weather station was downloaded from the same website as the three Woodstock stations. 

```{r}

library(tidyverse)
Crystal_Lake_temperature_data<-read_csv("~/Desktop/Ghatak_research_project/Crystal_Lake_Temperature_data.csv")
colnames(Crystal_Lake_temperature_data)<-c("Date", "Average_Temp_F", "Max_Temp_F", "Min_Temp_F", "PRCP_inches", "Snow_inches", "SNWD_inches")
Crystal_Lake_temperature_data<-Crystal_Lake_temperature_data%>%
tail(12582)%>%
mutate(Date=ymd(Date))%>%
mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
filter(Year==2025)%>%
tail(417)%>%
left_join(Averaged_LUREC_water_data, by=c("Date"="Date"))%>%
filter(Date<"2025-10-07" & Date>"2025-02-03")%>%
select(-2)%>%
mutate(Avg_Temp_F=(Max_Temp_F+Min_Temp_F)/2)


Combined_LUREC_water_data_new<-Combined_LUREC_water_data_pivoted%>%
mutate(Timestamp=mdy_hm(Timestamp))%>%
mutate(Day=format(Timestamp, format= "%d"), Month=format(Timestamp, format="%m"), Year=format(Timestamp, format="%y"))%>%
mutate(Date=str_c(Month, "/", Day, "/", Year))%>%
mutate(Date=mdy(Date))%>%
group_by(Date)%>%
summarize(Average_water_content=mean(Water_content))

Crystal_Lake_temperature_data<-Crystal_Lake_temperature_data%>%
left_join(Water_content_precipitation_combined, by=c("Date"="Date"))%>%
mutate(Avg_precip=(SW_0.7_Precip+SW_3.8_Precip+NW_5_Precip)/3)%>%
select(c(1,2,3,38,39))

colnames(Crystal_Lake_temperature_data)<-c("Date", "Max_Temp_F", "Min_Temp_F", "MWC", "Avg_precip")

Combined_Precip_water_content_max_temp_plot<-plot_ly(data=Crystal_Lake_temperature_data, x=~Date, y=~MWC, name="Mean Water Content(Left)", type="scatter", mode="lines+markers")%>%
  add_trace(x=~Date, y=~Avg_precip, yaxis="y2", name="Average Precip in Inches(Right)", type="bar")%>%  
  add_trace(x=~Date, y=~Max_Temp_F, yaxis="y3",  name="Crystal Lake 4 NW Max temp (Middle)", type="scatter", mode="lines+markers")%>%
  layout(title="Time Series of Mean Water Content, Maximum temperature and precipitation, averaged by day", xaxis=list(title="Date"), yaxis=list(title="Mean Water Content", side="left"), yaxis2=list(title="Crystal Lake precipitation", overlaying="y", side="right"), yaxis3=list(title="Maximum temperature", overlaying="y", position=0.5))
Combined_Precip_water_content_max_temp_plot


```

This graph suggested very rapid water infiltration, owing to the very small time delays ( usually <= 1 day) between precipitation events and increases in mean water content. It appears that, at certain points in the months of February and March, mean water content increases without precipitation, but this is partially due to missingness in the precipitation data during those months.

# Regression modeling

At this point, the relationship between maximum temperature and mean water content was uncertain and was investigated as part of the multiple linear regression model subsequently generated. The model regressed daily change in mean water content against daily maximum temperature from the Crystal Lake 4 NW weather station and the average of daily precipitation from Crystal Lake 4 NW, Woodstock 5 NW, and Woodstock 3.8 SW. 

First, the linearity assumption was assessed using a 3D scatterplot since two explanatory variables were involved. 

```{r}

Crystal_Lake_temperature_data<-Crystal_Lake_temperature_data%>%
mutate(Avg_temp_F=(Min_Temp_F+Max_Temp_F)/2)%>%
mutate(MWC_daily_change=MWC-lag(MWC))

Precip_plot<-plot_ly(Crystal_Lake_temperature_data, x=~Max_Temp_F, y=~Avg_precip, z=~MWC_daily_change)%>%
  add_markers()
Precip_plot

```

With the linearity assumption apparently satisfied, I generated the regression plane, checked the homoscedastictiy and residual normality assumptions, and then added the regression plane to the scatterplot. 

```{r}

MLR_model<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data)
summary(MLR_model)

plot(MLR_model, 1)
plot(MLR_model, 2)

```

Residuals generally appear scattered around zero. It is difficult to gauge if a fan shape is present because the precipitation data is very positively skewed. As I discuss later, this is one of the model's primary limitations.



There is a moderate violation of residual normality, with the normal qqplot appearing linear between -1 and 1 standard deviations of the mean of MWC_daily_change but exhibiting heavy tails on both ends. This might be alleviated by investigating the three potentially influential observations (observations 11, 77, and 121) using Jackknife residuals, testing if each observation is actually influential at alpha=0.01.


```{r}

x_seq<-seq(min(Crystal_Lake_temperature_data$Max_Temp_F, na.rm=TRUE), max(Crystal_Lake_temperature_data$Max_Temp_F, na.rm=TRUE), length.out=30)
y_seq<-seq(min(Crystal_Lake_temperature_data$Avg_precip, na.rm=TRUE), max(Crystal_Lake_temperature_data$Avg_precip, na.rm=TRUE), length.out=30)
grid<-expand.grid(Avg_precip=y_seq, Max_Temp_F=x_seq)
grid$MWC_daily_change<-predict(MLR_model, newdata=grid)
z_matrix<-matrix(grid$MWC_daily_change, nrow=length(y_seq), ncol=length(x_seq), byrow=FALSE)

Precip_plot <- plot_ly() %>%
  add_markers(data = Crystal_Lake_temperature_data,
              x = ~Max_Temp_F,
              y = ~Avg_precip,
              z = ~MWC_daily_change,
              marker = list(size = 3)) %>%
  add_surface(x = x_seq,
              y = y_seq,
              z = z_matrix,
              opacity = 0.6)

Precip_plot



```


```{r}

MWC_model_Jackknife_residuals<-rstudent(MLR_model)
Crystal_Lake_temperature_data$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model)))
Crystal_Lake_temperature_data$jackknife_resid[used_rows] <- MWC_model_Jackknife_residuals

n=nrow(Crystal_Lake_temperature_data)
k=3
p_value_11<-2*pt(abs(Crystal_Lake_temperature_data$jackknife_resid[11]), df=n-k-1, lower.tail=FALSE)

p_value_77<-2*pt(abs(Crystal_Lake_temperature_data$jackknife_resid[77]), df=n-k-1, lower.tail=FALSE)

p_value_121<-2*pt(abs(Crystal_Lake_temperature_data$jackknife_resid[121]), df=n-k-1, lower.tail=FALSE)

p_value_11
p_value_77
p_value_121



```
All three observations are statistically influential at alpha=0.01, so I will remove the observation with the highest absolute Jackknife residual value and regenerate the model without that observation.

```{r}

Crystal_Lake_temperature_data$jackknife_resid[11]
Crystal_Lake_temperature_data$jackknife_resid[77]
Crystal_Lake_temperature_data$jackknife_resid[121]

Crystal_Lake_temperature_data_2<-Crystal_Lake_temperature_data[-121, ]
new_MLR_model<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_2)
summary(new_MLR_model)
plot(new_MLR_model, 1)
plot(new_MLR_model, 2)




```
Removing the single influential observation did not significantly improve the residual normality. Nevertheless, I will continue testing and removing influential observations one at a time until none remain.

```{r}

Jackknife_residuals_2<-rstudent(new_MLR_model)

Crystal_Lake_temperature_data_2$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(new_MLR_model)))
Crystal_Lake_temperature_data_2$jackknife_resid[used_rows] <- Jackknife_residuals_2

n2=nrow(Crystal_Lake_temperature_data_2)
k2=3
p_value_11_new<-2*pt(abs(Crystal_Lake_temperature_data_2$jackknife_resid[11]), df=n2-k2-1, lower.tail=FALSE)
p_value_25_new<-2*pt(abs(Crystal_Lake_temperature_data_2$jackknife_resid[25]), df=n2-k2-1, lower.tail=FALSE)
p_value_77_new<-2*pt(abs(Crystal_Lake_temperature_data_2$jackknife_resid[77]), df=n2-k2-1, lower.tail=FALSE)

p_value_11_new
p_value_25_new
p_value_77_new

#All 3 potentially influential observations are statistically influential

Crystal_Lake_temperature_data_2$jackknife_resid[11]
Crystal_Lake_temperature_data_2$jackknife_resid[25]
Crystal_Lake_temperature_data_2$jackknife_resid[77]


```

Jackknife residual 11 from the dataset with one observation removed has the highest absolute residual value, so I will remove it and regenerate the model again.

```{r}

Crystal_Lake_temperature_data_3<-Crystal_Lake_temperature_data_2[-11, ]
MLR_model_v3<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_3)
summary(MLR_model_v3)
plot(MLR_model_v3, 1)
plot(MLR_model_v3, 2)

Jackknife_residuals_3<-rstudent(MLR_model_v3)

Crystal_Lake_temperature_data_3$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v3)))
Crystal_Lake_temperature_data_3$jackknife_resid[used_rows] <- Jackknife_residuals_3

n3=nrow(Crystal_Lake_temperature_data_3)
k3=3
p_value_10_new<-2*pt(abs(Crystal_Lake_temperature_data_3$jackknife_resid[10]), df=n3-k3-1, lower.tail=FALSE)
p_value_24_new<-2*pt(abs(Crystal_Lake_temperature_data_3$jackknife_resid[24]), df=n3-k3-1, lower.tail=FALSE)
p_value_76_new<-2*pt(abs(Crystal_Lake_temperature_data_3$jackknife_resid[76]), df=n3-k3-1, lower.tail=FALSE)

p_value_10_new
p_value_24_new
p_value_76_new

Crystal_Lake_temperature_data_3$jackknife_resid[10]
Crystal_Lake_temperature_data_3$jackknife_resid[24]
Crystal_Lake_temperature_data_3$jackknife_resid[76]


```
Observation 76 from the dataset with two influential observations removed has the highest absolute residual value, so I will remove it and regenerate the model for the third time.

```{r}

Crystal_Lake_temperature_data_4<-Crystal_Lake_temperature_data_3[-76, ]
MLR_model_v4<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_4)
summary(MLR_model_v4)
plot(MLR_model_v4, 1)
plot(MLR_model_v4, 2)

#Observations 10, 193, and 24 appear potentially influential.

Jackknife_residuals_4<-rstudent(MLR_model_v4)

Crystal_Lake_temperature_data_4$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v4)))
Crystal_Lake_temperature_data_4$jackknife_resid[used_rows] <- Jackknife_residuals_4

n4=nrow(Crystal_Lake_temperature_data_4)
k4=3
p_value_10_new<-2*pt(abs(Crystal_Lake_temperature_data_4$jackknife_resid[10]), df=n4-k4-1, lower.tail=FALSE)
p_value_193<-2*pt(abs(Crystal_Lake_temperature_data_4$jackknife_resid[193]), df=n4-k4-1, lower.tail=FALSE)
p_value_24_new<-2*pt(abs(Crystal_Lake_temperature_data_4$jackknife_resid[24]), df=n4-k4-1, lower.tail=FALSE)

p_value_10_new
p_value_193
p_value_24_new

Crystal_Lake_temperature_data_4$jackknife_resid[10]
Crystal_Lake_temperature_data_4$jackknife_resid[193]
Crystal_Lake_temperature_data_4$jackknife_resid[24]

```
All three observations are significantly influential, with observation 10 having the highest abolsute residual value. Thus, I will remove the 10th observation and regenerate the model a 4th time.

```{r}

Crystal_Lake_temperature_data_5<-Crystal_Lake_temperature_data_4[-10, ]
MLR_model_v5<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_5)
summary(MLR_model_v5)
plot(MLR_model_v5, 1)
plot(MLR_model_v5, 2)

#Observations 10, 192, and 23 appear potentially influential.

Jackknife_residuals_5<-rstudent(MLR_model_v5)

Crystal_Lake_temperature_data_5$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v5)))
Crystal_Lake_temperature_data_5$jackknife_resid[used_rows] <- Jackknife_residuals_5

n5=nrow(Crystal_Lake_temperature_data_5)
k5=3
p_value_10_new<-2*pt(abs(Crystal_Lake_temperature_data_5$jackknife_resid[10]), df=n5-k5-1, lower.tail=FALSE)
p_value_192<-2*pt(abs(Crystal_Lake_temperature_data_5$jackknife_resid[192]), df=n5-k5-1, lower.tail=FALSE)
p_value_23<-2*pt(abs(Crystal_Lake_temperature_data_5$jackknife_resid[23]), df=n5-k5-1, lower.tail=FALSE)

p_value_10_new
p_value_192
p_value_23

Crystal_Lake_temperature_data_5$jackknife_resid[10]
Crystal_Lake_temperature_data_5$jackknife_resid[192]
Crystal_Lake_temperature_data_5$jackknife_resid[24]

```
All 3 points are statistically influential, with observation 192 having the highest absolute residual value. I will regenerate the model a fifth time with observation 192 removed.

```{r}
Crystal_Lake_temperature_data_6<-Crystal_Lake_temperature_data_5[-192, ]
MLR_model_v6<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_6)
summary(MLR_model_v6)
plot(MLR_model_v6, 1)
plot(MLR_model_v6, 2)

#Observations 10, 23, and 24 appear potentially influential.

Jackknife_residuals_6<-rstudent(MLR_model_v6)

Crystal_Lake_temperature_data_6$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v6)))
Crystal_Lake_temperature_data_6$jackknife_resid[used_rows] <- Jackknife_residuals_6

n6=nrow(Crystal_Lake_temperature_data_6)
k6=3
p_value_10_new<-2*pt(abs(Crystal_Lake_temperature_data_6$jackknife_resid[10]), df=n6-k6-1, lower.tail=FALSE)
p_value_23<-2*pt(abs(Crystal_Lake_temperature_data_6$jackknife_resid[192]), df=n6-k6-1, lower.tail=FALSE)
p_value_24_new<-2*pt(abs(Crystal_Lake_temperature_data_6$jackknife_resid[23]), df=n6-k6-1, lower.tail=FALSE)

p_value_10_new
p_value_23
p_value_24_new

Crystal_Lake_temperature_data_6$jackknife_resid[10]
Crystal_Lake_temperature_data_6$jackknife_resid[23]
Crystal_Lake_temperature_data_6$jackknife_resid[24]


```
The 23rd observation is not statistically influential at alpha=0.01. Observations 10 and 24 are influential, and of these two data, observation 10 has the higher absolute residual value. Thus, I regenerate the model a sixth time, removing observation 10.

```{r}

Crystal_Lake_temperature_data_7<-Crystal_Lake_temperature_data_6[-10, ]
MLR_model_v7<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_7)
summary(MLR_model_v7)
plot(MLR_model_v7, 1)
plot(MLR_model_v7, 2)

#Observations 9, 22, and 23 appear potentially influential.

Jackknife_residuals_7<-rstudent(MLR_model_v7)

Crystal_Lake_temperature_data_7$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v7)))
Crystal_Lake_temperature_data_7$jackknife_resid[used_rows] <- Jackknife_residuals_7

n7=nrow(Crystal_Lake_temperature_data_7)
k7=3
p_value_9<-2*pt(abs(Crystal_Lake_temperature_data_7$jackknife_resid[9]), df=n7-k7-1, lower.tail=FALSE)
p_value_22<-2*pt(abs(Crystal_Lake_temperature_data_7$jackknife_resid[22]), df=n7-k7-1, lower.tail=FALSE)
p_value_23<-2*pt(abs(Crystal_Lake_temperature_data_7$jackknife_resid[23]), df=n7-k7-1, lower.tail=FALSE)

p_value_9
p_value_22
p_value_23

Crystal_Lake_temperature_data_7$jackknife_resid[9]
Crystal_Lake_temperature_data_7$jackknife_resid[22]
Crystal_Lake_temperature_data_7$jackknife_resid[23]

```
All 3 points are influential, with observation 22 having the highest abolsute residual value. I will regenerate the model a seventh time, removing observation 22.

```{r}

Crystal_Lake_temperature_data_8<-Crystal_Lake_temperature_data_7[-22, ]
MLR_model_v8<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_8)
summary(MLR_model_v8)
plot(MLR_model_v8, 1)
plot(MLR_model_v8, 2)

#Observations 9, 22, and 116 appear potentially influential.

Jackknife_residuals_8<-rstudent(MLR_model_v8)

Crystal_Lake_temperature_data_8$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v8)))
Crystal_Lake_temperature_data_8$jackknife_resid[used_rows] <- Jackknife_residuals_8

n8=nrow(Crystal_Lake_temperature_data_8)
k8=3
p_value_9<-2*pt(abs(Crystal_Lake_temperature_data_8$jackknife_resid[9]), df=n8-k8-1, lower.tail=FALSE)
p_value_22<-2*pt(abs(Crystal_Lake_temperature_data_8$jackknife_resid[22]), df=n8-k8-1, lower.tail=FALSE)
p_value_116<-2*pt(abs(Crystal_Lake_temperature_data_8$jackknife_resid[116]), df=n8-k8-1, lower.tail=FALSE)

p_value_9
p_value_22
p_value_116

Crystal_Lake_temperature_data_8$jackknife_resid[9]
Crystal_Lake_temperature_data_8$jackknife_resid[22]
Crystal_Lake_temperature_data_8$jackknife_resid[116]

```
```{r}
Crystal_Lake_temperature_data_9<-Crystal_Lake_temperature_data_8[-22, ]
MLR_model_v9<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_9)
summary(MLR_model_v9)
plot(MLR_model_v9, 1)
plot(MLR_model_v9, 2)

Jackknife_residuals_9<-rstudent(MLR_model_v9)

Crystal_Lake_temperature_data_9$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v9)))
Crystal_Lake_temperature_data_9$jackknife_resid[used_rows] <- Jackknife_residuals_9

n9=nrow(Crystal_Lake_temperature_data_9)
k9=3
p_value_9<-2*pt(abs(Crystal_Lake_temperature_data_9$jackknife_resid[9]), df=n9-k9-1, lower.tail=FALSE)
p_value_115<-2*pt(abs(Crystal_Lake_temperature_data_9$jackknife_resid[115]), df=n9-k9-1, lower.tail=FALSE)
p_value_117<-2*pt(abs(Crystal_Lake_temperature_data_9$jackknife_resid[117]), df=n9-k9-1, lower.tail=FALSE)

p_value_9
p_value_115
p_value_117

Crystal_Lake_temperature_data_9$jackknife_resid[9]
Crystal_Lake_temperature_data_9$jackknife_resid[115]
Crystal_Lake_temperature_data_9$jackknife_resid[117]

```
```{r}

Crystal_Lake_temperature_data_10<-Crystal_Lake_temperature_data_9[-115, ]
MLR_model_v10<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_10)
summary(MLR_model_v10)
plot(MLR_model_v10, 1)
plot(MLR_model_v10, 2)

Jackknife_residuals_10<-rstudent(MLR_model_v10)

Crystal_Lake_temperature_data_10$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v10)))
Crystal_Lake_temperature_data_10$jackknife_resid[used_rows] <- Jackknife_residuals_10

n10=nrow(Crystal_Lake_temperature_data_10)
k10=3
p_value_9<-2*pt(abs(Crystal_Lake_temperature_data_10$jackknife_resid[9]), df=n10-k10-1, lower.tail=FALSE)
p_value_115<-2*pt(abs(Crystal_Lake_temperature_data_10$jackknife_resid[115]), df=n10-k10-1, lower.tail=FALSE)
p_value_116<-2*pt(abs(Crystal_Lake_temperature_data_10$jackknife_resid[116]), df=n10-k10-1, lower.tail=FALSE)

p_value_9
p_value_115
p_value_116

Crystal_Lake_temperature_data_10$jackknife_resid[9]
Crystal_Lake_temperature_data_10$jackknife_resid[115]
Crystal_Lake_temperature_data_10$jackknife_resid[116]

```
```{r}

Crystal_Lake_temperature_data_11<-Crystal_Lake_temperature_data_10[-9, ]
MLR_model_v11<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_11)
summary(MLR_model_v11)
plot(MLR_model_v11, 1)
plot(MLR_model_v11, 2)

Jackknife_residuals_11<-rstudent(MLR_model_v11)

Crystal_Lake_temperature_data_11$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v11)))
Crystal_Lake_temperature_data_11$jackknife_resid[used_rows] <- Jackknife_residuals_11

n11=nrow(Crystal_Lake_temperature_data_11)
k11=3
p_value_114<-2*pt(abs(Crystal_Lake_temperature_data_11$jackknife_resid[114]), df=n11-k11-1, lower.tail=FALSE)
p_value_115<-2*pt(abs(Crystal_Lake_temperature_data_11$jackknife_resid[115]), df=n11-k11-1, lower.tail=FALSE)
p_value_180<-2*pt(abs(Crystal_Lake_temperature_data_11$jackknife_resid[180]), df=n11-k11-1, lower.tail=FALSE)

p_value_114
p_value_115
p_value_180

Crystal_Lake_temperature_data_11$jackknife_resid[114]
Crystal_Lake_temperature_data_11$jackknife_resid[115]
Crystal_Lake_temperature_data_11$jackknife_resid[180]

```
After removing 10 influential observation, I have reached the point where only 1 influential observation remains: observation 115. I will regenerate the model for the eleventh time, removing observation 115.

```{r}

Crystal_Lake_temperature_data_12<-Crystal_Lake_temperature_data_11[-115, ]
MLR_model_v12<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_12)
summary(MLR_model_v12)
plot(MLR_model_v12, 1)
plot(MLR_model_v12, 2)

Jackknife_residuals_12<-rstudent(MLR_model_v12)

Crystal_Lake_temperature_data_12$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v12)))
Crystal_Lake_temperature_data_12$jackknife_resid[used_rows] <- Jackknife_residuals_12

n12=nrow(Crystal_Lake_temperature_data_12)
k12=3
p_value_100<-2*pt(abs(Crystal_Lake_temperature_data_12$jackknife_resid[100]), df=n12-k12-1, lower.tail=FALSE)
p_value_114<-2*pt(abs(Crystal_Lake_temperature_data_12$jackknife_resid[114]), df=n12-k12-1, lower.tail=FALSE)
p_value_179<-2*pt(abs(Crystal_Lake_temperature_data_12$jackknife_resid[179]), df=n12-k12-1, lower.tail=FALSE)

p_value_100
p_value_114
p_value_179

Crystal_Lake_temperature_data_12$jackknife_resid[100]
Crystal_Lake_temperature_data_12$jackknife_resid[114]
Crystal_Lake_temperature_data_12$jackknife_resid[179]

```
We are back to three influential observations; however, for the sake of space, I will stop the Jackknifing process here and use MLR_model_v12 for inference and predictions.

```{r}

summary(MLR_model_v12)

```
Interpretation of model:

* 73.86% of the sample variation in MWC_daily_change is explained through linear relationships with Max_Temp_F and Avg_precip.

* Based on the sample, we expect MWC_daily_change to decrease by 4.964*10^-5 m^3 water/m^3 soil per 1 degree Fahrenheit increase in Max_Temp_F, holding Avg_precip constant. This physically makes sense since higher daily temperature yields higher evaporation rates, thereby yielding more dramatic decreases in soil water content.

* We expect MWC_daily_change to increase by 0.01718 m^3 water/m^3 soil per 1-inch increase in Avg_precip, holding Max_Temp_F constant. This physically makes sense since precipitation increases soil moisture. 

* The model has a significantly better overall fitting performance than the null regression plane at the sample mean of MWC_daily_change (as indicated by the small p value of the overall F test).

* Both predictors significantly improve the fitting performance of the model including only the other predictor, as indicated by the small p values of the t tests for those individual predictors (which are statistically equivalent to single partial F tests between the respective reduced models and the full model containing both predictors). 


An important point to note with this model:

Before removing any influential observations, the p value for the t test statistic associated with the Max_Temp_F variable was 0.8372, meaning that it did not significantly improve the fitting performance of the model already including the Avg_precip variable.
However, after removing some influential observations, this t test statistic became significant at alpha=0.05, taking on a value of 0.00231. This means that, when influential observations are accounted for, Max_Temp_F does significantly contribute to the model's fitting performance. When I originally created this regression model for my research poster, I did not do any Jackknifing as I had not learned it yet in my Regression Analysis class. When I did learn about jackknife residuals, I was occupied with writing the proposal for future research and did not think to remove influential observations from the model. Thus, my poster states that Max_Temp_F is not significant. 

Thankfully, this was a preliminary model created from temperature and precipitation data collected from proxy sources. Future research will utilize on-site temperature sensors and precipitation gauges to eliminate uncertainty associated with spatial variability. It was thus a low-stakes opportunity to learn an important lesson that I will carry with me for the rest of my career: ALWAYS check for influential observations every time I run a regression model. Not only do influential observations affect parameter estimates, they can also dramatically change the p values of partial significance hypothesis tests, affecting a researcher's decision making. 

Moving on to test for multicollinearity:

```{r}

VIF(MLR_model_v12)

```
Variance inflation factors are very low (close to 1) for both predictors, yielding no evidence of multicollinearity. Mathematically, this means that the model regressing Max_Temp_F on Avg_precip and the model regressing Avg_precip on Max_Temp_F both have very low R squared values, since $VIF(j)=1/(1-Rj^2)$. 

Moving on to inference using the model. Remember that the response variable is MWC_daily_change, meaning the model predicts the increase or decrease in mean water content between the previous day and the current day. Let's choose two random consecutive observations from the data and run the model on the more recent observation.

```{r}

obs49<-Crystal_Lake_temperature_data_12[49, ]
obs50<-Crystal_Lake_temperature_data_12[50, ]
MWC_daily_change_actual<-obs50$MWC-obs49$MWC
obs49
obs50
MWC_daily_change_actual

```
Under Max_Temp_F=73 degrees F and Avg_precip=0.2433333, a change in MWC of 0.00144 was observed. Let's see what the model predicts will happen to MWC under these conditions.

```{r}

newdata<-data.frame(Max_Temp_F=73, Avg_precip=0.2433333)
predict(MLR_model_v12, newdata, interval="prediction", level=0.95)

```
The model predicts a 1-day change in MWC of about 0.00109 m^3 water/m^3 of soil for a single day with these atmospheric conditions, which is similar to the actual observed value.

After soil cover materials and onsite temperature sensors and precipitation gauges are installed, the future regression model will include the following variables:

* WC_daily_change (continuous): the response variable, measuring how much the average daily water content for a given pot has changed between the previous day and the current day

* pot_no (categorical): the number of the pot (1-12) as previously labeled

* daily_precip (continuous): the raw onsite measurement of daily precipitation as recorded by the precipitation gauge

* daily_max_temp_F (continuous): the onsite measurement of daily maximum temperature as recorded by the temperature sensor

* cover_material_type (categorical): the cover material on top of each pot (permeable tile, cement, vegetation, or nothing)

#personal access token: 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