---
title: "LUREC_research_code"
author: "Justin Peabody"
date: "2026-04-30"
output: md_document
---

```{r}

library(tidyverse)
library(plotly)
library(regclass)

```

The very first action taken, before the project began, was downloading nearby precipitation data from this site: <https://www.ncei.noaa.gov/access/past-weather/Woodstock%2C%20Illinois.>

This was later visualized against the soil water content data from the field site. 

```{r, echo=FALSE}

Woodstock_0.7_SW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_0.7_SW_weather_data.csv")
Woodstock_3.8_SW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_3.8_SW_data.csv")
Woodstock_5_NW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_5_NW_data.csv")

```
I then prepared the data for visualization and analysis by reformatting each date field to date class, filtering rows for 2025, and combining each station's data into a single dataset.

```{r}

Woodstock_0.7_SW_data<-Woodstock_0.7_SW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))

Woodstock_3.8_SW_data<-Woodstock_3.8_SW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))
  
Woodstock_5_NW_data<-Woodstock_5_NW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))
  
Woodstock_combined_weather_data<-Woodstock_0.7_SW_data%>%
  left_join(Woodstock_3.8_SW_data, by=c("Date"="Date"))%>%
  left_join(Woodstock_5_NW_data, by=c("Date"="Date"))%>%
  select(1,2,6,10)

names(Woodstock_combined_weather_data)<-c("Date","SW_0.7_Precip", "SW_3.8_Precip", "NW_5_Precip")

```

At the onset of my Environmental Research class in January, I obtained the onsite soil moisture data from my professor, which includes moisture data from November 10, 2024 (just after being installed) to October 6, 2025.

```{r}
six_LUREC_water_sensor_data<-read_csv("~/Desktop/Ghatak_research_project/6_LUREC_soil_water_sensors.csv")
six__more_LUREC_water_sensors<-read_csv("~/Desktop/Ghatak_research_project/6_more_LUREC_water_sensors.csv")

```
Each dataset contains readings from a single data logger, which receives data from six of the twelve sensors (2 datasets = 2 data loggers = 12 total sensors). 

I then performed some basic data cleaning, including removal of column headers that were formatted as observations and renaming columns to prevent word splitting issues, and changing water content readings from character to numeric class.

```{r}

six_LUREC_water_sensor_data_new<-six_LUREC_water_sensor_data%>%
tail(7916)
colnames(six_LUREC_water_sensor_data_new)<-c("Timestamp", "Pot_1_Water_Content_proportion", "Pot_2_Water_Content_proportion", "Pot_3_Water_Content_proportion", "Pot_4_Water_Content_proportion", "Pot_5_Water_Content_proportion", "Pot_6_Water_Content_proportion", "Battery_percent", "Battery_voltage_mV")

six__more_LUREC_water_sensors_new<-six__more_LUREC_water_sensors%>%
tail(7910)
colnames(six__more_LUREC_water_sensors_new)<-c("Timestamp", "Pot_7_Water_Content_proportion", "Pot_8_Water_Content_proportion", "Pot_9_Water_Content_proportion", "Pot_10_Water_Content_proportion", "Pot_11_Water_Content_proportion", "Pot_12_Water_Content_proportion", "Battery_percent", "Battery_voltage_mV")

rm(six_LUREC_water_sensor_data, six__more_LUREC_water_sensors)

six_LUREC_water_sensor_data_new<-six_LUREC_water_sensor_data_new%>%
mutate(Pot_1_Water_Content_proportion=as.numeric(Pot_1_Water_Content_proportion))%>%
mutate(Pot_2_Water_Content_proportion=as.numeric(Pot_2_Water_Content_proportion))%>%
mutate(Pot_3_Water_Content_proportion=as.numeric(Pot_3_Water_Content_proportion))%>%
mutate(Pot_4_Water_Content_proportion=as.numeric(Pot_4_Water_Content_proportion))%>%
mutate(Pot_5_Water_Content_proportion=as.numeric(Pot_5_Water_Content_proportion))%>%
mutate(Pot_6_Water_Content_proportion=as.numeric(Pot_6_Water_Content_proportion))

six_more_LUREC_water_sensors_new<-six__more_LUREC_water_sensors_new%>%
  mutate(Pot_7_Water_Content_proportion=as.numeric(Pot_7_Water_Content_proportion))%>%
  mutate(Pot_8_Water_Content_proportion=as.numeric(Pot_8_Water_Content_proportion))%>%
  mutate(Pot_9_Water_Content_proportion=as.numeric(Pot_9_Water_Content_proportion))%>%
  mutate(Pot_10_Water_Content_proportion=as.numeric(Pot_10_Water_Content_proportion))%>%
  mutate(Pot_11_Water_Content_proportion=as.numeric(Pot_11_Water_Content_proportion))%>%
  mutate(Pot_12_Water_Content_proportion=as.numeric(Pot_12_Water_Content_proportion))

rm(six__more_LUREC_water_sensors_new)

six_LUREC_water_sensor_data_new<-six_LUREC_water_sensor_data_new%>%
tail(7915)
Combined_LUREC_water_data<-six_LUREC_water_sensor_data_new%>%
left_join(six_more_LUREC_water_sensors_new, by=c("Timestamp"="Timestamp"))

```

I then investigated the distribution of each pot's water content.

```{r}

Averaged_LUREC_water_data<-Combined_LUREC_water_data%>%
mutate(Timestamp=mdy_hm(Timestamp))%>%
mutate(Day=format(Timestamp, format= "%d"), Month=format(Timestamp, format="%m"), Year=format(Timestamp, format="%y"))%>%
mutate(Date=str_c(Month, "/", Day, "/", Year))%>%
mutate(Date=mdy(Date))%>%
group_by(Date)%>%
summarize(Pot_1_mean_water_content=mean(Pot_1_Water_Content_proportion), Pot_2_mean_water_content=mean(Pot_2_Water_Content_proportion), Pot_3_mean_water_content=mean(Pot_3_Water_Content_proportion), Pot_4_mean_water_content=mean(Pot_4_Water_Content_proportion), Pot_5_mean_water_content=mean(Pot_5_Water_Content_proportion), Pot_6_mean_water_content=mean(Pot_6_Water_Content_proportion), Pot_7_mean_water_content=mean(Pot_7_Water_Content_proportion), Pot_8_mean_water_content=mean(Pot_8_Water_Content_proportion), Pot_9_mean_water_content=mean(Pot_9_Water_Content_proportion), Pot_10_mean_water_content=mean(Pot_10_Water_Content_proportion), Pot_11_mean_water_content=mean(Pot_11_Water_Content_proportion), Pot_12_mean_water_content=mean(Pot_12_Water_Content_proportion))    

Water_content_precipitation_combined<-Averaged_LUREC_water_data%>%
left_join(Woodstock_combined_weather_data, by=c("Date"="Date"))%>%
mutate(Avg_MWC=(Pot_1_mean_water_content+Pot_2_mean_water_content+Pot_3_mean_water_content+Pot_4_mean_water_content+Pot_5_mean_water_content+Pot_6_mean_water_content+Pot_7_mean_water_content+Pot_8_mean_water_content+Pot_9_mean_water_content+Pot_10_mean_water_content+Pot_11_mean_water_content+Pot_12_mean_water_content)/12)%>%
select(-17)

colnames(Water_content_precipitation_combined)=c("Date", "Pot_1_MWC", "Pot_2_MWC", "Pot_3_MWC", "Pot_4_MWC", "Pot_5_MWC", "Pot_6_MWC", "Pot_7_MWC", "Pot_8_MWC", "Pot_9_MWC", "Pot_10_MWC", "Pot_11_MWC", "Pot_12_MWC","PRCP_x", "PRCP_y", "PRCP_z")
Water_content_precipitation_combined<-Water_content_precipitation_combined%>%
mutate(Avg_precip=(PRCP_x+PRCP_y+PRCP_z)/3)

Combined_LUREC_water_data<-six_LUREC_water_sensor_data_new%>%
tail(7915)%>%
left_join(six_more_LUREC_water_sensors_new, by=c("Timestamp"="Timestamp"))

Combined_LUREC_water_data_pivoted<-Combined_LUREC_water_data%>%
select(-Battery_percent.x, -Battery_voltage_mV.x, -Battery_percent.y, -Battery_voltage_mV.y)%>%
pivot_longer(cols=-Timestamp, names_to="Pot", values_to="Water_content")

Combined_LUREC_water_data_pivoted$Pot<-rep(1:12, 7915)
Combined_LUREC_water_data_pivoted$Pot<-as.factor(Combined_LUREC_water_data_pivoted$Pot)

Pot_water_content_boxplots<-plot_ly(data=Combined_LUREC_water_data_pivoted, color=~Pot, y=~Water_content, type="box")%>%
layout(title="Comparative boxplots of Water Content for Each LUREC Pot", xaxis=list(title="Pot Number"), yaxis=list(title="Water content (m^3 water/m^3 soil"))
Pot_water_content_boxplots

```

Creating this boxplot made it visually apparent that pots 5 and 8 contained negative values. Since soil moisture content is measured as a proportion of cubic meters of water per cubic meter of soil, negative moisture readings make no sense. 

I then investigated when these negative readings occured.

```{r}

Pot_5_negative_values<-Combined_LUREC_water_data%>%
select(Timestamp, Pot_5_Water_Content_proportion)%>%
filter(Pot_5_Water_Content_proportion<0)

Pot_8_negative_values<-Combined_LUREC_water_data%>%
select(Timestamp, Pot_8_Water_Content_proportion)%>%
filter(Pot_8_Water_Content_proportion<0)

```

The sensor in Pot 5 recorded negative values for water content at each hour mark from 1/21/25 at 9:00 AM to 1/22/25 at 7:00 PM. The sensor in Pot 8 recorded negative values for water content at each hour mark from 10:00 AM to 5:00 PM on 1/21/26. The negative values then discontinue and begin again from 12:00 AM to 5:00 PM on 1/22/25.

Based on this information and the locations of large clusters of NA values between the months of November and January for 2 of the pots, I decided to remove all observations at or before 2/3/25, after averaging the observations by day. This was the dataset used for all subsequent analyses.

```{r}

Averaged_LUREC_water_data<-Combined_LUREC_water_data%>%
mutate(Timestamp=mdy_hm(Timestamp))%>%
mutate(Day=format(Timestamp, format= "%d"), Month=format(Timestamp, format="%m"), Year=format(Timestamp, format="%y"))%>%
mutate(Date=str_c(Month, "/", Day, "/", Year))%>%
mutate(Date=mdy(Date))%>%
group_by(Date)%>%
summarize(Pot_1_mean_water_content=mean(Pot_1_Water_Content_proportion), Pot_2_mean_water_content=mean(Pot_2_Water_Content_proportion), Pot_3_mean_water_content=mean(Pot_3_Water_Content_proportion), Pot_4_mean_water_content=mean(Pot_4_Water_Content_proportion), Pot_5_mean_water_content=mean(Pot_5_Water_Content_proportion), Pot_6_mean_water_content=mean(Pot_6_Water_Content_proportion), Pot_7_mean_water_content=mean(Pot_7_Water_Content_proportion), Pot_8_mean_water_content=mean(Pot_8_Water_Content_proportion), Pot_9_mean_water_content=mean(Pot_9_Water_Content_proportion), Pot_10_mean_water_content=mean(Pot_10_Water_Content_proportion), Pot_11_mean_water_content=mean(Pot_11_Water_Content_proportion), Pot_12_mean_water_content=mean(Pot_12_Water_Content_proportion)) 

Water_content_precipitation_combined<-Averaged_LUREC_water_data%>%
left_join(Woodstock_combined_weather_data, by=c("Date"="Date"))%>%
mutate(Avg_MWC=(Pot_1_mean_water_content+Pot_2_mean_water_content+Pot_3_mean_water_content+Pot_4_mean_water_content+Pot_5_mean_water_content+Pot_6_mean_water_content+Pot_7_mean_water_content+Pot_8_mean_water_content+Pot_9_mean_water_content+Pot_10_mean_water_content+Pot_11_mean_water_content+Pot_12_mean_water_content)/12)

Useful_LUREC_averaged_data<-Water_content_precipitation_combined%>%
filter(Date > "2025-02-03")

```

I then created a three-axis time series graph displaying daily averaged water content against the daily precipitation and maximum temperature at the Crystal Lake 4 NW weather station (the only weather station near the field site with both temperature and precipitation data for 2025). Data for this weather station was downloaded from the same website as the three Woodstock stations. 

```{r}

library(tidyverse)
Crystal_Lake_temperature_data<-read_csv("~/Desktop/Ghatak_research_project/Crystal_Lake_Temperature_data.csv")
colnames(Crystal_Lake_temperature_data)<-c("Date", "Average_Temp_F", "Max_Temp_F", "Min_Temp_F", "PRCP_inches", "Snow_inches", "SNWD_inches")
Crystal_Lake_temperature_data<-Crystal_Lake_temperature_data%>%
tail(12582)%>%
mutate(Date=ymd(Date))%>%
mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
filter(Year==2025)%>%
tail(417)%>%
left_join(Averaged_LUREC_water_data, by=c("Date"="Date"))%>%
filter(Date<"2025-10-07" & Date>"2025-02-03")%>%
select(-2)%>%
mutate(Avg_Temp_F=(Max_Temp_F+Min_Temp_F)/2)


Combined_LUREC_water_data_new<-Combined_LUREC_water_data_pivoted%>%
mutate(Timestamp=mdy_hm(Timestamp))%>%
mutate(Day=format(Timestamp, format= "%d"), Month=format(Timestamp, format="%m"), Year=format(Timestamp, format="%y"))%>%
mutate(Date=str_c(Month, "/", Day, "/", Year))%>%
mutate(Date=mdy(Date))%>%
group_by(Date)%>%
summarize(Average_water_content=mean(Water_content))

Crystal_Lake_temperature_data<-Crystal_Lake_temperature_data%>%
left_join(Water_content_precipitation_combined, by=c("Date"="Date"))%>%
mutate(Avg_precip=(SW_0.7_Precip+SW_3.8_Precip+NW_5_Precip)/3)%>%
select(c(1,2,3,38,39))

colnames(Crystal_Lake_temperature_data)<-c("Date", "Max_Temp_F", "Min_Temp_F", "MWC", "Avg_precip")

Combined_Precip_water_content_max_temp_plot<-plot_ly(data=Crystal_Lake_temperature_data, x=~Date, y=~MWC, name="Mean Water Content(Left)", type="scatter", mode="lines+markers")%>%
  add_trace(x=~Date, y=~Avg_precip, yaxis="y2", name="Average Precip in Inches(Right)", type="bar")%>%  
  add_trace(x=~Date, y=~Max_Temp_F, yaxis="y3",  name="Crystal Lake 4 NW Max temp (Middle)", type="scatter", mode="lines+markers")%>%
  layout(title="Time Series of Mean Water Content, Maximum temperature and precipitation, averaged by day", xaxis=list(title="Date"), yaxis=list(title="Mean Water Content", side="left"), yaxis2=list(title="Crystal Lake precipitation", overlaying="y", side="right"), yaxis3=list(title="Maximum temperature", overlaying="y", position=0.5))
Combined_Precip_water_content_max_temp_plot


```

This graph suggested very rapid water infiltration, owing to the very small time delays ( usually <= 1 day) between precipitation events and increases in mean water content. It appears that, at certain points in the months of February and March, mean water content increases without precipitation, but this is partially due to missingness in the precipitation data during those months. 

At this point, the relationship between maximum temperature and mean water content was uncertain and was investigated as part of the multiple linear regression model subsequently generated. The model regressed daily change in mean water content against daily maximum temperature from the Crystal Lake 4 NW weather station and the average of daily precipitation from Crystal Lake 4 NW, Woodstock 5 NW, and Woodstock 3.8 SW. 

First, the linearity assumption was assessed using a 3D scatterplot since two explanatory variables were involved. 

```{r}

Crystal_Lake_temperature_data<-Crystal_Lake_temperature_data%>%
mutate(Avg_temp_F=(Min_Temp_F+Max_Temp_F)/2)%>%
mutate(MWC_daily_change=MWC-lag(MWC))

Precip_plot<-plot_ly(Crystal_Lake_temperature_data, x=~Max_Temp_F, y=~Avg_precip, z=~MWC_daily_change)%>%
  add_markers()
Precip_plot

```

With the linearity assumption apparently satisfied, I generated the regression plane, checked the homoscedastictiy and residual normality assumptions, and then added the regression plane to the scatterplot. 

```{r}

MLR_model<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data)
summary(MLR_model)

plot(MLR_model, 1)
plot(MLR_model, 2)

```
Residuals generally appear scattered around zero. It is difficult to gauge if a fan shape is present because the precipitation data is very positively skewed. As I discuss later, this is one of the model's primary limitations.

```{r}

x_seq<-seq(min(Crystal_Lake_temperature_data$Max_Temp_F, na.rm=TRUE), max(Crystal_Lake_temperature_data$Max_Temp_F, na.rm=TRUE), length.out=30)
y_seq<-seq(min(Crystal_Lake_temperature_data$Avg_precip, na.rm=TRUE), max(Crystal_Lake_temperature_data$Avg_precip, na.rm=TRUE), length.out=30)
grid<-expand.grid(Avg_precip=y_seq, Max_Temp_F=x_seq)
grid$MWC_daily_change<-predict(MLR_model, newdata=grid)
z_matrix<-matrix(grid$MWC_daily_change, nrow=length(y_seq), ncol=length(x_seq), byrow=FALSE)

Precip_plot <- plot_ly() %>%
  add_markers(data = Crystal_Lake_temperature_data,
              x = ~Max_Temp_F,
              y = ~Avg_precip,
              z = ~MWC_daily_change,
              marker = list(size = 3)) %>%
  add_surface(x = x_seq,
              y = y_seq,
              z = z_matrix,
              opacity = 0.6)

Precip_plot



```


There is a moderate violation of residual normality, with the normal qqplot appearing linear between -1 and 1 standard deviations of the mean of MWC_daily_change but exhibiting heavy tails on both ends. This might be alleviated by investigating the three potentially influential observations (observations 11, 77, and 121) using Jackknife residuals, testing if each observation is actually influential at alpha=0.01.

```{r}

MWC_model_Jackknife_residuals<-rstudent(MLR_model)
Crystal_Lake_temperature_data$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model)))
Crystal_Lake_temperature_data$jackknife_resid[used_rows] <- MWC_model_Jackknife_residuals

n=nrow(Crystal_Lake_temperature_data)
k=3
p_value_11<-2*pt(abs(Crystal_Lake_temperature_data$jackknife_resid[11]), df=n-k-1, lower.tail=FALSE)

p_value_77<-2*pt(abs(Crystal_Lake_temperature_data$jackknife_resid[77]), df=n-k-1, lower.tail=FALSE)

p_value_121<-2*pt(abs(Crystal_Lake_temperature_data$jackknife_resid[121]), df=n-k-1, lower.tail=FALSE)

p_value_11
p_value_77
p_value_121



```
All three observations are statistically influential at alpha=0.01, so I will remove the observation with the highest absolute Jackknife residual value and regenerate the model without that observation.

```{r}

Crystal_Lake_temperature_data$jackknife_resid[11]
Crystal_Lake_temperature_data$jackknife_resid[77]
Crystal_Lake_temperature_data$jackknife_resid[121]

Crystal_Lake_temperature_data_2<-Crystal_Lake_temperature_data[-121, ]
new_MLR_model<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_2)
summary(new_MLR_model)
plot(new_MLR_model, 1)
plot(new_MLR_model, 2)




```
Removing the single influential observation did not singificantly improve the residual normality. Nevertheless, I will continue testing and removing influential observations one at a time until none remain.

```{r}

Jackknife_residuals_2<-rstudent(new_MLR_model)

Crystal_Lake_temperature_data_2$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(new_MLR_model)))
Crystal_Lake_temperature_data_2$jackknife_resid[used_rows] <- Jackknife_residuals_2

n2=nrow(Crystal_Lake_temperature_data_2)
k2=3
p_value_11_new<-2*pt(abs(Crystal_Lake_temperature_data_2$jackknife_resid[11]), df=n2-k2-1, lower.tail=FALSE)
p_value_25_new<-2*pt(abs(Crystal_Lake_temperature_data_2$jackknife_resid[25]), df=n2-k2-1, lower.tail=FALSE)
p_value_77_new<-2*pt(abs(Crystal_Lake_temperature_data_2$jackknife_resid[77]), df=n2-k2-1, lower.tail=FALSE)

p_value_11_new
p_value_25_new
p_value_77_new

#All 3 potnetially influential observations are statistically influential

Crystal_Lake_temperature_data_2$jackknife_resid[11]
Crystal_Lake_temperature_data_2$jackknife_resid[25]
Crystal_Lake_temperature_data_2$jackknife_resid[77]


```

Jackknife residual 11 from the dataset with one observation removed has the highest absolute residual value, so I will remove it and regenerate the model again.

```{r}

Crystal_Lake_temperature_data_3<-Crystal_Lake_temperature_data_2[-11, ]
MLR_model_v3<-lm(MWC_daily_change~Max_Temp_F+Avg_precip, data=Crystal_Lake_temperature_data_3)
summary(MLR_model_v3)
plot(MLR_model_v3, 1)
plot(MLR_model_v3, 2)

Jackknife_residuals_3<-rstudent(MLR_model_v3)

Crystal_Lake_temperature_data_3$jackknife_resid <- NA
used_rows <- as.numeric(rownames(model.frame(MLR_model_v3)))
Crystal_Lake_temperature_data_3$jackknife_resid[used_rows] <- Jackknife_residuals_3

n3=nrow(Crystal_Lake_temperature_data_3)
k3=3
p_value_10_new<-2*pt(abs(Crystal_Lake_temperature_data_3$jackknife_resid[10]), df=n3-k3-1, lower.tail=FALSE)
p_value_24_new<-2*pt(abs(Crystal_Lake_temperature_data_3$jackknife_resid[24]), df=n3-k3-1, lower.tail=FALSE)
p_value_76_new<-2*pt(abs(Crystal_Lake_temperature_data_3$jackknife_resid[76]), df=n3-k3-1, lower.tail=FALSE)

p_value_10_new
p_value_24_new
p_value_76_new

Crystal_Lake_temperature_data_3$jackknife_resid[10]
Crystal_Lake_temperature_data_3$jackknife_resid[24]
Crystal_Lake_temperature_data_3$jackknife_resid[76]


```
Observation 76 from the dataset with two influential observations removed has the highest abolsute residual value, so I will remove it and regenerate the model for the third time.