---
title: "LUREC_research_code"
author: "Justin Peabody"
date: "2026-04-30"
output: md_document
---

```{r}

library(tidyverse)
library(plotly)
```

The very first action taken, before the project began, was downloading nearby precipitation data from this site: <https://www.ncei.noaa.gov/access/past-weather/Woodstock%2C%20Illinois.>

This was later visualized against the soil water content data from the field site. 

```{r, echo=FALSE}

Woodstock_0.7_SW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_0.7_SW_weather_data.csv")
Woodstock_3.8_SW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_3.8_SW_data.csv")
Woodstock_5_NW_data<-read_csv("~/Desktop/Ghatak_research_project/Woodstock_5_NW_data.csv")

```
I then prepared the data for visualization and analysis by reformatting each date field to date class, filtering rows for 2025, and combining each station's data into a single dataset.

```{r}

Woodstock_0.7_SW_data<-Woodstock_0.7_SW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))

Woodstock_3.8_SW_data<-Woodstock_3.8_SW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))
  
Woodstock_5_NW_data<-Woodstock_5_NW_data%>%
  mutate(Date=ymd(Date))%>%
  mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
  filter(Year==2025)%>%
  select(c(1,5,8,9,10))
  
Woodstock_combined_weather_data<-Woodstock_0.7_SW_data%>%
  left_join(Woodstock_3.8_SW_data, by=c("Date"="Date"))%>%
  left_join(Woodstock_5_NW_data, by=c("Date"="Date"))%>%
  select(1,2,6,10)

names(Woodstock_combined_weather_data)<-c("Date","SW_0.7_Precip", "SW_3.8_Precip", "NW_5_Precip")

```

At the onset of my Environmental Research class in January, I obtained the onsite soil moisture data from my professor, which includes moisture data from November 10, 2024 (just after being installed) to October 6, 2025.

```{r}
six_LUREC_water_sensor_data<-read_csv("~/Desktop/Ghatak_research_project/6_LUREC_soil_water_sensors.csv")
six__more_LUREC_water_sensors<-read_csv("~/Desktop/Ghatak_research_project/6_more_LUREC_water_sensors.csv")

```
Each dataset contains readings from a single data logger, which receives data from six of the twelve sensors (2 datasets = 2 data loggers = 12 total sensors). 

I then performed some basic data cleaning, including removal of column headers that were formatted as observations and renaming columns to prevent word splitting issues, and changing water content readings from character to numeric class.

```{r}

six_LUREC_water_sensor_data_new<-six_LUREC_water_sensor_data%>%
tail(7916)
colnames(six_LUREC_water_sensor_data_new)<-c("Timestamp", "Pot_1_Water_Content_proportion", "Pot_2_Water_Content_proportion", "Pot_3_Water_Content_proportion", "Pot_4_Water_Content_proportion", "Pot_5_Water_Content_proportion", "Pot_6_Water_Content_proportion", "Battery_percent", "Battery_voltage_mV")

six__more_LUREC_water_sensors_new<-six__more_LUREC_water_sensors%>%
tail(7910)
colnames(six__more_LUREC_water_sensors_new)<-c("Timestamp", "Pot_7_Water_Content_proportion", "Pot_8_Water_Content_proportion", "Pot_9_Water_Content_proportion", "Pot_10_Water_Content_proportion", "Pot_11_Water_Content_proportion", "Pot_12_Water_Content_proportion", "Battery_percent", "Battery_voltage_mV")

rm(six_LUREC_water_sensor_data, six__more_LUREC_water_sensors)

six_LUREC_water_sensor_data_new<-six_LUREC_water_sensor_data_new%>%
mutate(Pot_1_Water_Content_proportion=as.numeric(Pot_1_Water_Content_proportion))%>%
mutate(Pot_2_Water_Content_proportion=as.numeric(Pot_2_Water_Content_proportion))%>%
mutate(Pot_3_Water_Content_proportion=as.numeric(Pot_3_Water_Content_proportion))%>%
mutate(Pot_4_Water_Content_proportion=as.numeric(Pot_4_Water_Content_proportion))%>%
mutate(Pot_5_Water_Content_proportion=as.numeric(Pot_5_Water_Content_proportion))%>%
mutate(Pot_6_Water_Content_proportion=as.numeric(Pot_6_Water_Content_proportion))

six_more_LUREC_water_sensors_new<-six__more_LUREC_water_sensors_new%>%
  mutate(Pot_7_Water_Content_proportion=as.numeric(Pot_7_Water_Content_proportion))%>%
  mutate(Pot_8_Water_Content_proportion=as.numeric(Pot_8_Water_Content_proportion))%>%
  mutate(Pot_9_Water_Content_proportion=as.numeric(Pot_9_Water_Content_proportion))%>%
  mutate(Pot_10_Water_Content_proportion=as.numeric(Pot_10_Water_Content_proportion))%>%
  mutate(Pot_11_Water_Content_proportion=as.numeric(Pot_11_Water_Content_proportion))%>%
  mutate(Pot_12_Water_Content_proportion=as.numeric(Pot_12_Water_Content_proportion))

rm(six__more_LUREC_water_sensors_new)

```

I then investigated the distribution of each pot's water content.

```{r}

Water_content_precipitation_combined<-Averaged_LUREC_water_data%>%
left_join(Woodstock_combined_weather_data, by=c("Date"="Date"))%>%
mutate(Avg_MWC=(Pot_1_mean_water_content+Pot_2_mean_water_content+Pot_3_mean_water_content+Pot_4_mean_water_content+Pot_5_mean_water_content+Pot_6_mean_water_content+Pot_7_mean_water_content+Pot_8_mean_water_content+Pot_9_mean_water_content+Pot_10_mean_water_content+Pot_11_mean_water_content+Pot_12_mean_water_content)/12)%>%
select(-17)

colnames(Water_content_precipitation_combined)=c("Date", "Pot_1_MWC", "Pot_2_MWC", "Pot_3_MWC", "Pot_4_MWC", "Pot_5_MWC", "Pot_6_MWC", "Pot_7_MWC", "Pot_8_MWC", "Pot_9_MWC", "Pot_10_MWC", "Pot_11_MWC", "Pot_12_MWC","PRCP_x", "PRCP_y", "PRCP_z")
Water_content_precipitation_combined<-Water_content_precipitation_combined%>%
mutate(Avg_precip=(PRCP_x+PRCP_y+PRCP_z)/3)

Combined_LUREC_water_data<-six_LUREC_water_sensor_data_new%>%
tail(7915)%>%
left_join(six_more_LUREC_water_sensors_new, by=c("Timestamp"="Timestamp"))

Combined_LUREC_water_data_pivoted<-Combined_LUREC_water_data%>%
select(-Battery_percent.x, -Battery_voltage_mV.x, -Battery_percent.y, -Battery_voltage_mV.y)%>%
pivot_longer(cols=-Timestamp, names_to="Pot", values_to="Water_content")

Combined_LUREC_water_data_pivoted$Pot<-rep(1:12, 7915)
Combined_LUREC_water_data_pivoted$Pot<-as.factor(Combined_LUREC_water_data_pivoted$Pot)

Pot_water_content_boxplots<-plot_ly(data=Combined_LUREC_water_data_pivoted, color=~Pot, y=~Water_content, type="box")%>%
layout(title="Comparative boxplots of Water Content for Each LUREC Pot", xaxis=list(title="Pot Number"), yaxis=list(title="Water content (m^3 water/m^3 soil"))
Pot_water_content_boxplots

```

Creating this boxplot made it visually apparent that pots 5 and 8 contained negative values. Since soil moisture content is measured as a proportion of cubic meters of water per cubic meter of soil, negative moisture readings make no sense. 

I then investigated when these negative readings occured.

```{r}

Pot_5_negative_values<-Combined_LUREC_water_data%>%
select(Timestamp, Pot_5_Water_Content_proportion)%>%
filter(Pot_5_Water_Content_proportion<0)

Pot_8_negative_values<-Combined_LUREC_water_data%>%
select(Timestamp, Pot_8_Water_Content_proportion)%>%
filter(Pot_8_Water_Content_proportion<0)

```

The sensor in Pot 5 recorded negative values for water content at each hour mark from 1/21/25 at 9:00 AM to 1/22/25 at 7:00 PM. The sensor in Pot 8 recorded negative values for water content at each hour mark from 10:00 AM to 5:00 PM on 1/21/26. The negative values then discontinue and begin again from 12:00 AM to 5:00 PM on 1/22/25.

Based on this information and the locations of large clusters of NA values between the months of November and January for 2 of the pots, I decided to remove all observations at or before 2/3/25, after averaging the observations by day. This was the dataset used for all subsequent analyses.

```{r}

Averaged_LUREC_water_data<-Combined_LUREC_water_data%>%
mutate(Timestamp=mdy_hm(Timestamp))%>%
mutate(Day=format(Timestamp, format= "%d"), Month=format(Timestamp, format="%m"), Year=format(Timestamp, format="%y"))%>%
mutate(Date=str_c(Month, "/", Day, "/", Year))%>%
mutate(Date=mdy(Date))%>%
group_by(Date)%>%
summarize(Pot_1_mean_water_content=mean(Pot_1_Water_Content_proportion), Pot_2_mean_water_content=mean(Pot_2_Water_Content_proportion), Pot_3_mean_water_content=mean(Pot_3_Water_Content_proportion), Pot_4_mean_water_content=mean(Pot_4_Water_Content_proportion), Pot_5_mean_water_content=mean(Pot_5_Water_Content_proportion), Pot_6_mean_water_content=mean(Pot_6_Water_Content_proportion), Pot_7_mean_water_content=mean(Pot_7_Water_Content_proportion), Pot_8_mean_water_content=mean(Pot_8_Water_Content_proportion), Pot_9_mean_water_content=mean(Pot_9_Water_Content_proportion), Pot_10_mean_water_content=mean(Pot_10_Water_Content_proportion), Pot_11_mean_water_content=mean(Pot_11_Water_Content_proportion), Pot_12_mean_water_content=mean(Pot_12_Water_Content_proportion)) 

Water_content_precipitation_combined<-Averaged_LUREC_water_data%>%
left_join(Woodstock_combined_weather_data, by=c("Date"="Date"))%>%
mutate(Avg_MWC=(Pot_1_mean_water_content+Pot_2_mean_water_content+Pot_3_mean_water_content+Pot_4_mean_water_content+Pot_5_mean_water_content+Pot_6_mean_water_content+Pot_7_mean_water_content+Pot_8_mean_water_content+Pot_9_mean_water_content+Pot_10_mean_water_content+Pot_11_mean_water_content+Pot_12_mean_water_content)/12)

Useful_LUREC_averaged_data<-Water_content_precipitation_combined%>%
filter(Date > "2025-02-03")

```

I then created a three-axis time series graph displaying daily averaged water content against the daily precipitation and maximum temperature at the Crystal Lake 4 NW weather station (the only weather station near the field site with both temperature and precipitation data for 2025). Data for this weather station was downloaded from the same website as the three Woodstock stations. 

```{r}

library(tidyverse)
Crystal_Lake_temperature_data<-read_csv("~/Desktop/Ghatak_research_project/Crystal_Lake_Temperature_data.csv")
colnames(Crystal_Lake_temperature_data)<-c("Date", "Average_Temp_F", "Max_Temp_F", "Min_Temp_F", "PRCP_inches", "Snow_inches", "SNWD_inches")
Crystal_Lake_temperature_data<-Crystal_Lake_temperature_data%>%
tail(12582)%>%
mutate(Date=ymd(Date))%>%
mutate(Year=year(Date), Month=month(Date), Day=day(Date))%>%
filter(Year==2025)%>%
tail(417)%>%
left_join(Averaged_LUREC_water_data, by=c("Date"="Date"))%>%
filter(Date<"2025-10-07" & Date>"2025-02-03")%>%
select(-2)%>%
mutate(Avg_Temp_F=(Max_Temp_F+Min_Temp_F)/2)


Combined_LUREC_water_data_new<-Combined_LUREC_water_data_pivoted%>%
mutate(Timestamp=mdy_hm(Timestamp))%>%
mutate(Day=format(Timestamp, format= "%d"), Month=format(Timestamp, format="%m"), Year=format(Timestamp, format="%y"))%>%
mutate(Date=str_c(Month, "/", Day, "/", Year))%>%
mutate(Date=mdy(Date))%>%
group_by(Date)%>%
summarize(Average_water_content=mean(Water_content))

Crystal_Lake_temperature_data<-Crystal_Lake_temperature_data%>%
left_join(Water_content_precipitation_combined, by=c("Date"="Date"))%>%
mutate(Avg_precip=(SW_0.7_Precip+SW_3.8_Precip+NW_5_Precip)/3)%>%
select(c(1,2,3,38,39))

colnames(Crystal_Lake_temperature_data)<-c("Date", "Max_Temp_F", "Min_Temp_F", "MWC", "Avg_precip")

Combined_Precip_water_content_max_temp_plot<-plot_ly(data=Crystal_Lake_temperature_data, x=~Date, y=~MWC, name="Mean Water Content(Left)", type="scatter", mode="lines+markers")%>%
  add_trace(x=~Date, y=~Avg_precip, yaxis="y2", name="Average Precip in Inches(Right)", type="bar")%>%  
  add_trace(x=~Date, y=~Max_Temp_F, yaxis="y3",  name="Crystal Lake 4 NW Max temp (Middle)", type="scatter", mode="lines+markers")%>%
  layout(title="Time Series of Mean Water Content, Maximum temperature and precipitation, averaged by day", xaxis=list(title="Date"), yaxis=list(title="Mean Water Content", side="left"), yaxis2=list(title="Crystal Lake precipitation", overlaying="y", side="right"), yaxis3=list(title="Maximum temperature", overlaying="y", position=0.5))
Combined_Precip_water_content_max_temp_plot


```



